Equalizer\\Programming and User Guide

### 2013-01-12 Developer/ProgrammingGuide/paper.tex
Fix typos

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/processes.pdf}
  {\caption{\label{fProcesses}Equalizer Processes}}
\end{wrapfigure}
Collage is a cross-platform C++ library for building heterogenous, distributed
applications. Collage provides an abstraction of different network connections,
peer-to-peer messaging, discovery and synchronization as well as
high-performance, object-oriented, versioned data distribution. Collage is
designed for low-overhead multi-threaded execution which allows applications to
easily exploit multi-core architectures. Equalizer uses Collage as the cluster
backend, e.g., by setting up direct communication between two nodes when needed
for image compositing or software swap barriers.

\fig{fProcesses} depicts the relationship between the server, application,
render client and adminstrative processes, which are explained below.

@@

Equalizer 1.0 introduced the admin library, which can be used to modify a
running Equalizer server. The admin library is still in development, but already
allows limited modifications such as adding new windows and changing
layouts. The admin library may be used to create standalone administration tools
or from within the application code. In any case, it has an indepedent view of
the server's configuration. Documentation for admin library is not yet part of
this Programming and User Guide.

@@

\begin{itemize}
\item The default value is \textsf{local} which uses all local GPUs queried
  using the cgl, glx or wgl GPU-SD module.
\item \textsf{--eq-config sessionname} uses the dns\_sd ZeroConf module of
  GPU-SD to query all GPUs in the subnet for the given session. The default
  session of the GPU-SD daemon is \textsf{default}. The hostname provided by
  ZeroConf is used to connect the nodes.
\item \textsf{--eq-config filename.eqc} loads a static configuration from the
  given ascii file. The following sections describe how to write configuration
  files.
\end{itemize}

@@

\begin{itemize}
\item A list of all \textbf{computers} in your rendering cluster,
  including the \textbf{IP addresses} of all network interfaces to be
  used.
\item The number of \textbf{graphics cards} in each computer.
\item The \textbf{physical dimensions} of the display system, if
  applicable. These are typically the bottom-left, bottom-right and
  top-left corner points of each display surface in meters.
\item The relative coordinates of all the \textbf{segments} belonging to
  each display surface, and the graphics card \textbf{output} used for
  each segment. For homogenous setups, it is often enough to know the
  number of rows and columns on each surface, as well as the overlap or
  underlap percentage, if applicable.
\item The number of desired \textbf{application windows}. Application
  windows are typically destination windows for scalable rendering or
  'control' windows paired with a view on a display system.
\item \textbf{Characteristics} of the application, e.g., supported
  scalability modes and features.
\end{itemize}

@@

The \textsf{eq::Window} class encapsulates a drawable and an OpenGL
context. The drawable can be an on-screen window or an off-screen
PBuffer or FrameBuffer Object (FBO).

@@

When using hardware swap synchronization, use swap barriers for all segment to
be synchronized, setting \textsf{NV\_group} and \textsf{NV\_barrier}
appropriately. The swap barrier name is ignored in this case. All windows of the
same group on a single node synchronize their swap buffer. All groups of the
same barrier synchronize their swap buffer across nodes. Please note that the
driver typically limits the number of groups and barriers to one, and that
multiple windows per swap group are not supported by all drivers. Hardware swap
barriers require support from the OpenGL driver, and has been tested on nVidia
Quadro GPUs with the G-Sync option. Please refer to your OpenGL driver
documentation for details.

@@

A software swap barrier is configured by giving it a name. Windows with a swap
barrier of the same name synchronize with each other before executing the swap
buffer task. Before entering the barrier, \textsf{Window::finish} is called to
ensure that all OpenGL commands have been executed.  Swap barrier names are only
valid within the compound tree, that is, a compound from one compound tree
cannot be synchronized with a compound from another ompound tree.

@@

Display walls made out of LCD's, monoscopic or passive stereo projection systems
often only use a software swap barrier and no framelock. The display bezels make
it very hard to notice the missing synchronization.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
  {\caption{\label{fViewLoadBalancing}\small Cross-Segment
      Load-Balancing for two Segments using eight GPU's}}
\end{wrapfigure}
\fig{fViewLoadBalancing}\footnote{Image Copyright Realtime Technology
  AG, 2008} illustrates this process. On the left side, a static
assignment of resources to display segments is used. The right-hand
segment has a higher load than the left-hand segment, causing
sub-optimal performance. The configuration on the right uses a view
equalizer, which assigns two GPU's to the left segment and four GPU's to
the right segment, which leads to optimal performance for this model and
camera position.

@@

\begin{wrapfigure}{l}{.382\textwidth}
    \includegraphics[width=.382\textwidth]{images/viewLB.png}
    {\caption{\label{fViewLoadBalancing2}\small Cross-Segment
        Load-Balancing for a CAVE}}
\end{wrapfigure}
\fig{fViewLoadBalancing2} shows cross-usage for a five-sided CAVE driven
by five GPU's. The front and left segments show the model and have a
significant rendering load. The view equalizer assigns the GPU's from
the top, bottom and right wall for rendering the left and front wall in
this configuration.

@@

\subsection{Name Resolution}
Pay attention that your name resolution works properly, that is, the output of
the hostname command should resolve to the same, non-local IP address on all
machines. If this is not the case, you will have to provide the public IP
address for all processes, in the following way:
\begin{description}
\item[Server] Specify the server IP as the hostname field in the connection
  description in the server section.
\item[Application] Specify the application IP as the hostname field in the
  connection description in the appNode section. If the application should not
  contribute to the rendering, set up an appNode without a pipe section.
\item[Render Clients] Specify the client IP's as the hostname field in the
  connection description of each node section.
\end{description}

@@

\subsection{Debugging}
If your configuration does not work, simplify it as much as possible
first. Normally this means that there is one server, application and render
client. Failure to launch a cluster configuration often is caused by one of the
following reasons:
\begin{itemize}
\item A firewall is blocking network connections.
\item The render client can't access the GPU's on the remote host. Set up your
  X-Server or application rights correctly. Log into the remote machine using
  the launch command and try to execute a simple application, e.g.,
  \textsf{glxinfo -display :0.1}.
\item The server does not find the prelaunched render client. Verify that the
  client is listening on the correct IP and port, and that this IP and port are
  reachable from the server host.
\item The server cannot launch a render client. Check the server log for the
  launch command used, and try to execute a simple application from the server
  host using this launch command. It should run without user interaction. Check
  that the render client is installed in the correct path. Pay attention to the
  launch command quotes used to separate arguments on Windows. Check that the
  same software versions, including Equalizer, are installed on all machines.
\item A client can't connect back to the application. Check the client log, this
  is typically caused by a misconfigured host name resolution.
\end{itemize}

@@

\begin{description}
\item[seq] The core namespace for Sequel, the simple interface to the Equalizer
  client library.
\item[eq] The core namespace for the Equalizer client library. The classes and
  their relationship in this namespace closely model the configuration file
  format. The classes in the \textsf{eq} namespace are the main subject of this
  Programming Guide.\fig{fClientUml} provides an overview map of the most
  important classes in the Equalizer namespace, grouped by functionality.
\item[eq::util] The eq::util namespace provides common utility classes, which
  often simplify the usage of OpenGL functions. Most classes in this namespace
  are used by the Equalizer client library, but are usable independently
  from Equalizer for application development.
\item[eq::admin] The eq::admin namespace implements an administrative API to
  change the configurations of a running server. This admin API is not yet
  finalized and will very likely change in the future.
\item[eq::fabric] The eq::fabric namespace is the common data management and
  transport layer between the client library, the server and the administrative
  API. Most Equalizer client classes inherit from base classes in this namespace
  and have all their data access methods in these base classes.
\item[co] Collage is the network library used by Equalizer. It provides basic
  functionality for network communication, such as \textsf{Connection} and
  \textsf{ConnectionSet}, as well as higher-level functionality such as
  \textsf{Node}, \textsf{LocalNode}, \textsf{Object} and
  \textsf{Serializable}. Please refer to \sref{sNetwork} for an introduction
  into the network layer, and to \sref{sNetObject} for distributed objects.
\item[lunchbox] The lunchbox library provides C++ classes to abstract the
  underlying operating system and to implement common helper functionality for
  multi-threaded applications. Examples are \textsf{lunchbox::Clock} providing a
  high-resolution timer, or \textsf{lunchbox::MTQueue} providing a thread-safe,
  blocking fifo. Classes in this namespace are fully documented in the API
  documentation on the Equalizer website, and are not subject of this
  Programming Guide.
\item[hwloc, boost, vmmlib, gpu-sd] External libraries providing manual and
  automatic thread affinity, serialization and the foundation for RSP multicast,
  vector and matrix mathematics as well as local and remote GPU discovery,
  respectively. The hwloc and gpu-sd libraries are used only internally and not
  exposed through the API.
\item[eq::server] The server namespace, implementing the functionality of the
  Equalizer server, which is an internal namespace not to be used by application
  developers. The \textsf{eq::admin} namespace enables run-time administration
  of Equalizer servers. The server does not yet expose a stable
  API.
\end{description}

@@

Custom data type serializers can be implemented by providing the
appropriate serialization functions. No pointers should be directly
transmitted through the data streams. For pointers, the corresponding
object is typically a distributed object as well, and it's identifier
and potentially version is transmitted in place of its pointer.

@@

First, it configures the objects to retain its data \textsf{latency+1} commits,
which corresponds to the typical use case where objects are commited once per
frame. This allows render clients, which often are behind the application
process, to map objects with an old version. This does not necessarily translate
into increased memory usage, since new versions are only created when the object
was dirty during commit.

@@

The event handling code repdroduced here is just showing the handling of one
type of event. A detailed description on how to customize event handling can be
found in \sref{sEventHandling}:

@@

{\footnotesize\begin{lstlisting}
void eq::Channel::frameDraw( const uin128_t& frameID )
{
    applyBuffer();
    applyViewport();

@@

The actual rendering uses display lists or vertex buffer objects. These
OpenGL objects are allocated using the object manager. The rendering is
done by the leaf nodes, which are small enough to store the vertex
indices in a \textsf{short} value for optimal performance with VBO's.
The leaf nodes reuse the objects stored in the object manager, or create
and set up new objects if it was not yet set up. Since one object
manager is used per thread (pipe), this allows a thread-safe sharing of
the compiled display lists or VBO's across all windows of a pipe.

@@

Carbon events issued for AGL windows are initially receive by the node
main thread, and automatically dispatched to the pipe thread. The pipe
thread will dispatch the event to the appropriate event handler. One
\textsf{AGLEventHandler} per window is used, which is created during
\textsf{AGLWindow::configInit}. The event handler installs a Carbon
event handler for all important events. The event handler uses an
\textsf{AGLWindowEvent} to pass the Carbon \textsf{EventRef} to
\textsf{AGLWindowIF::processEvent}. During window exit, the installed
Carbon handler is removed when the window's event handler is deleted. No
event handling is set up for PBuffers.

For each GLX window, one \textsf{GLXEventHandler} is allocated. FBO's and
PBuffers are handled in the same way as window drawables. The event dispatch
finds the corresponding event handler for each received event, which is then
used to process the event in a fashion similar to AGL. The
\textsf{GLXWindowEvent} passes the \textsf{XEvent} to
\textsf{GLXWindowIF::\-process\-Event}.

Each \textsf{WGLWindow} allocates one \textsf{WGLEventHandler} when the
window handle is set. The WGLEventHandler passes the native event
parameters \textsf{uMsg}, \textsf{wParam} and \textsf{lParam} to
\textsf{WGL\-Win\-dowIF::processEvent} as part of the
\textsf{WGLWindowEvent}. No event handling is set up for PBuffers.

@@

This section explains window initialization in detail. It discusses in
detail the handling of the different window systems. Window systems are
integrated through the \textsf{SystemWindow} interface, which externalizes
the details of the windowing API from the \textsf{eq::Window}
implementation and facilitates the integration with new windowing API's
and custom applications.

@@

An application typically choses to subclass the sample implementation if only
minor tweaks are needed for integration. For major changes or new window
systems, it is often easier to subclass directly from \textsf{SystemWindow} or
\textsf{GLWindow} and implement the abstract methods of this interface class.

@@

\begin{enumerate}
\item\textsf{initWGLAffinityDC} is used to set up an affinity device
  context, which might be needed for window creation. The WGL window
  tracks potentially two device context handles, one for OpenGL context
  creation (the affinity DC), and one for \textsf{swapBuffers} (the
  window's DC).
\item\textsf{chooseWGLPixelFormat} chooses a pixel format based on the
  window attributes. If no device context is given, it uses the system
  device context. The chosen pixel format is set on the passed device
  context.
\item\textsf{configInitWGLDrawable} creates the drawable. The device
  context passed to \textsf{configInitWGLDrawable} is used to query the
  pixel format and is used as the device context for creating a
  PBuffer. If no device context is given, the display device context is
  used. On success, it sets the window handle. Setting a window handle
  also sets the window's device context.
\item\textsf{createWGLContext} creates an OpenGL rendering context using
  the given device context. If no device context is given, the window's
  device context is used. This function does not set the window's OpenGL
  context.
\end{enumerate}

@@

For head-mounted displays (HMD), the tracking information is used to move the
frusta with the observer, as shown in \fig{fHMD}. This results in different
projections compared to normal tracking with fixed projection screens. This
difference is transparent to Equalizer applications, only the configuration file
uses a different wall type for HMD's.

@@

The focus distance calculation relative to the observer is similar to the origin
algorithm, but it keeps the closest wall in the observer's view direction at the
given focus distance, as shown in \fig{fObserver}. When the observer moves
forward, the focal plane moves forward as well. Consequently, when the observer
looks in a different direction, a different object in the scene is focused, as
indictated by the dotted circle in \fig{fObserver}.

@@

On a high level, the compositor combines multiple input frames using 2D
tiling, depth-compositing for polygonal data or sorted, alpha-blended
compositing for semitransparent volumetric data. These operations
composite either directly all images on the GPU, or use a CPU-based
compositor and then transfer the preintegrated result to the GPU. The
high-level entry points automatically select the best algorithm. The
CPU-based compositor uses, based on availability, OpenMP and the
Paracomp compositing library to accelerate its operation.

@@

Each \textsf{LocalNode} provides a \textsf{co::Zeroconf} communicator, which
allows node and resource discovery. This requires compiling using a Lunchbox
library supporting dns\_sd. The service "\_collage.\_tcp" is used to announce
the presence of a listening \textsf{LocalNode} using the zeroconf protocol to
the network, unless the local node has no listening connections. The node
identifier and all listening connection descriptions are announced using keys
starting with "co\_". Internally Collage uses this information to connect
unknown nodes by using the node identifier alone.

Applications may use the zeroconf communicator to add additional key-value pairs
to announce application-specific data, and to retrieve a snapshot of all
key-value pairs of all discovered nodes on the network.

@@

\textsf{co::Serializable} implements one typical usage pattern for data
distribution bsed on \textsf{co::Object}. The \textsf{co::Serializable} data
distribution is founded on the concept of dirty bits, allowing inheritance with
data distribution. Dirty bits form a 64-bit mask which marks which parts of the
object have to be distributed during the next commit. It is easier to use, but
imposes one typical way to implement data distribution.

@@

\begin{tabularx}{\textwidth}{|l|X|}
  \hline
  \textbf{Name} & EQ\_CONFIG\_IATTR\_FOCUS\_MODE\\
  \textbf{Value} & fixed \textbar relative\_to\_origin \textbar
  relative\_to\_observer\\
  \textbf{Default} & fixed\\
  \textbf{Details} & The mode how to use the focus distance. The
  focus maode and distance influence the calculation of stereo frusta. See
  \sref{sFocusDistance} for details.\\
  \textbf{See also} & EQ\_CONFIG\_IATTR\_FOCUS\_DISTANCE
  EQ\_CONFIG\_FATTR\_EYE\_BASE\\
  \hline
\end{tabularx}\\\vfill

@@

\begin{tabularx}{\textwidth}{|l|X|}
  \hline
  \textbf{Name} & EQ\_WINDOW\_IATTR\_HINT\_STEREO\\
  \textbf{Value} & OFF \textbar \ ON \textbar \ AUTO\\
  \textbf{Default} & AUTO\\
  \textbf{Details} & Determines if the window selects a quad-buffered
  stereo visual. When set to AUTO, the default window initialization methods try
  to allocate a stereo visual for windows, but fall back to a mono
  visual if allocation fails. For PBuffers, AUTO selects a mono visual.\\
  \textbf{See also} & EQ\_COMPOUND\_IATTR\_STEREO\_MODE\\
  \hline
\end{tabularx}\\\vfill

\begin{tabularx}{\textwidth}{|l|X|}
  \hline
  \textbf{Name} & EQ\_WINDOW\_IATTR\_HINT\_DOUBLEBUFFER\\
  \textbf{Value} & OFF \textbar \ ON \textbar \ AUTO\\
  \textbf{Default} & AUTO\\
  \textbf{Details} & Determines if the window selects a double-buffered
  stereo visual. When set to AUTO, the default window initialization methods try
  to allocate a double-buffered visual for windows, but fall back to a
  single-buffered visual if allocation fails. For PBuffers, AUTO selects
  a single-buffered visual.\\
  \hline
\end{tabularx}\\\vfill

@@

\begin{tabularx}{\textwidth}{|l|X|}
  \hline
  \textbf{Name} & EQ\_WINDOW\_IATTR\_HINT\_DRAWABLE\\
  \textbf{Value} & window \textbar \ pbuffer \textbar \ FBO \textbar \ OFF\\
  \textbf{Default} & window\\
  \textbf{Details} & Selects the window's drawable type. A window is an
  on-screen, window system-dependent window with a full-window OpenGL
  drawable. PBuffers are off-screen drawables created using window
  system-dependent PBuffer API's. FBO are off-screen frame buffer
  objects. A disabled drawable creates a system window without a frame buffer
  for context sharing, e.g., for asynchronous operations in a separate
  thread. To calculate the PBuffer or FBO size on unconnected devices, a pipe
  viewport size of 4096x4096 is assumed, unless specified otherwise using the
  pipe's viewport parameter.\\
  \hline
\end{tabularx}\\\vfill

>>>>

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/processes.pdf}
  {\caption{\label{fProcesses}Equalizer Processes}}
\end{wrapfigure}
Collage is a cross-platform C++ library for building heterogeneous, distributed
applications. Collage provides an abstraction of different network connections,
peer-to-peer messaging, discovery and synchronization as well as
high-performance, object-oriented, versioned data distribution. Collage is
designed for low-overhead multi-threaded execution which allows applications to
easily exploit multi-core architectures. Equalizer uses Collage as the cluster
backend, e.g., by setting up direct communication between two nodes when needed
for image compositing or software swap barriers.

\fig{fProcesses} depicts the relationship between the server, application,
render client and administrative processes, which are explained below.

@@

Equalizer 1.0 introduced the admin library, which can be used to modify a
running Equalizer server. The admin library is still in development, but already
allows limited modifications such as adding new windows and changing
layouts. The admin library may be used to create standalone administration tools
or from within the application code. In any case, it has an independent view of
the server's configuration. Documentation for admin library is not yet part of
this Programming and User Guide.

@@

\begin{itemize}
\item The default value is \textsf{local} which uses all local GPUs queried
  using the cgl, glx or wgl GPU-SD module.
\item \textsf{--eq-config sessionname} uses the dns\_sd ZeroConf module of
  GPU-SD to query all GPUs in the subnet for the given session. The default
  session of the GPU-SD daemon is \textsf{default}. The hostname provided by
  ZeroConf is used to connect the nodes.
\item \textsf{--eq-config filename.eqc} loads a static configuration from the
  given ASCII file. The following sections describe how to write configuration
  files.
\end{itemize}

@@

\begin{itemize}
\item A list of all \textbf{computers} in your rendering cluster,
  including the \textbf{IP addresses} of all network interfaces to be
  used.
\item The number of \textbf{graphics cards} in each computer.
\item The \textbf{physical dimensions} of the display system, if
  applicable. These are typically the bottom-left, bottom-right and
  top-left corner points of each display surface in meters.
\item The relative coordinates of all the \textbf{segments} belonging to
  each display surface, and the graphics card \textbf{output} used for
  each segment. For homogeneous setups, it is often enough to know the
  number of rows and columns on each surface, as well as the overlap or
  underlap percentage, if applicable.
\item The number of desired \textbf{application windows}. Application
  windows are typically destination windows for scalable rendering or
  'control' windows paired with a view on a display system.
\item \textbf{Characteristics} of the application, e.g., supported
  scalability modes and features.
\end{itemize}

@@

The \textsf{eq::Window} class encapsulates a drawable and an OpenGL
context. The drawable can be an on-screen window or an off-screen
pbuffer or framebuffer Object (FBO).

@@

When using hardware swap synchronization, use swap barriers for all segment to
be synchronized, setting \textsf{NV\_group} and \textsf{NV\_barrier}
appropriately. The swap barrier name is ignored in this case. All windows of the
same group on a single node synchronize their swap buffer. All groups of the
same barrier synchronize their swap buffer across nodes. Please note that the
driver typically limits the number of groups and barriers to one, and that
multiple windows per swap group are not supported by all drivers. Hardware swap
barriers require support from the OpenGL driver, and has been tested on NVIDIA
Quadro GPUs with the G-Sync option. Please refer to your OpenGL driver
documentation for details.

@@

A software swap barrier is configured by giving it a name. Windows with a swap
barrier of the same name synchronize with each other before executing the swap
buffer task. Before entering the barrier, \textsf{Window::finish} is called to
ensure that all OpenGL commands have been executed.  Swap barrier names are only
valid within the compound tree, that is, a compound from one compound tree
cannot be synchronized with a compound from another compound tree.

@@

Display walls made out of LCDs, monoscopic or passive stereo projection systems
often only use a software swap barrier and no framelock. The display bezels make
it very hard to notice the missing synchronization.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
  {\caption{\label{fViewLoadBalancing}\small Cross-Segment
      Load-Balancing for two Segments using eight GPUs}}
\end{wrapfigure}
\fig{fViewLoadBalancing}\footnote{Image Copyright Realtime Technology
  AG, 2008} illustrates this process. On the left side, a static
assignment of resources to display segments is used. The right-hand
segment has a higher load than the left-hand segment, causing
sub-optimal performance. The configuration on the right uses a view
equalizer, which assigns two GPUs to the left segment and four GPUs to
the right segment, which leads to optimal performance for this model and
camera position.

@@

\begin{wrapfigure}{l}{.382\textwidth}
    \includegraphics[width=.382\textwidth]{images/viewLB.png}
    {\caption{\label{fViewLoadBalancing2}\small Cross-Segment
        Load-Balancing for a CAVE}}
\end{wrapfigure}
\fig{fViewLoadBalancing2} shows cross-usage for a five-sided CAVE driven
by five GPUs. The front and left segments show the model and have a
significant rendering load. The view equalizer assigns the GPUs from
the top, bottom and right wall for rendering the left and front wall in
this configuration.

@@

\subsection{Name Resolution}
Pay attention that your name resolution works properly, that is, the output of
the hostname command should resolve to the same, non-local IP address on all
machines. If this is not the case, you will have to provide the public IP
address for all processes, in the following way:
\begin{description}
\item[Server] Specify the server IP as the hostname field in the connection
  description in the server section.
\item[Application] Specify the application IP as the hostname field in the
  connection description in the appNode section. If the application should not
  contribute to the rendering, set up an appNode without a pipe section.
\item[Render Clients] Specify the client IPs as the hostname field in the
  connection description of each node section.
\end{description}

@@

\subsection{Debugging}
If your configuration does not work, simplify it as much as possible
first. Normally this means that there is one server, application and render
client. Failure to launch a cluster configuration often is caused by one of the
following reasons:
\begin{itemize}
\item A firewall is blocking network connections.
\item The render client can't access the GPUs on the remote host. Set up your
  X-Server or application rights correctly. Log into the remote machine using
  the launch command and try to execute a simple application, e.g.,
  \textsf{glxinfo -display :0.1}.
\item The server does not find the prelaunched render client. Verify that the
  client is listening on the correct IP and port, and that this IP and port are
  reachable from the server host.
\item The server cannot launch a render client. Check the server log for the
  launch command used, and try to execute a simple application from the server
  host using this launch command. It should run without user interaction. Check
  that the render client is installed in the correct path. Pay attention to the
  launch command quotes used to separate arguments on Windows. Check that the
  same software versions, including Equalizer, are installed on all machines.
\item A client can't connect back to the application. Check the client log, this
  is typically caused by a misconfigured host name resolution.
\end{itemize}

@@

\begin{description}
\item[seq] The core namespace for Sequel, the simple interface to the Equalizer
  client library.
\item[eq] The core namespace for the Equalizer client library. The classes and
  their relationship in this namespace closely model the configuration file
  format. The classes in the \textsf{eq} namespace are the main subject of this
  Programming Guide.\fig{fClientUml} provides an overview map of the most
  important classes in the Equalizer namespace, grouped by functionality.
\item[eq::util] The eq::util namespace provides common utility classes, which
  often simplify the usage of OpenGL functions. Most classes in this namespace
  are used by the Equalizer client library, but are usable independently
  from Equalizer for application development.
\item[eq::admin] The eq::admin namespace implements an administrative API to
  change the configurations of a running server. This admin API is not yet
  finalized and will very likely change in the future.
\item[eq::fabric] The eq::fabric namespace is the common data management and
  transport layer between the client library, the server and the administrative
  API. Most Equalizer client classes inherit from base classes in this namespace
  and have all their data access methods in these base classes.
\item[co] Collage is the network library used by Equalizer. It provides basic
  functionality for network communication, such as \textsf{Connection} and
  \textsf{ConnectionSet}, as well as higher-level functionality such as
  \textsf{Node}, \textsf{LocalNode}, \textsf{Object} and
  \textsf{Serializable}. Please refer to \sref{sNetwork} for an introduction
  into the network layer, and to \sref{sNetObject} for distributed objects.
\item[lunchbox] The lunchbox library provides C++ classes to abstract the
  underlying operating system and to implement common helper functionality for
  multi-threaded applications. Examples are \textsf{lunchbox::Clock} providing a
  high-resolution timer, or \textsf{lunchbox::MTQueue} providing a thread-safe,
  blocking FIFO. Classes in this namespace are fully documented in the API
  documentation on the Equalizer website, and are not subject of this
  Programming Guide.
\item[hwloc, boost, vmmlib, gpu-sd] External libraries providing manual and
  automatic thread affinity, serialization and the foundation for RSP multicast,
  vector and matrix mathematics as well as local and remote GPU discovery,
  respectively. The hwloc and gpu-sd libraries are used only internally and not
  exposed through the API.
\item[eq::server] The server namespace, implementing the functionality of the
  Equalizer server, which is an internal namespace not to be used by application
  developers. The \textsf{eq::admin} namespace enables run-time administration
  of Equalizer servers. The server does not yet expose a stable
  API.
\end{description}

@@

Custom data type serializers can be implemented by providing the
appropriate serialization functions. No pointers should be directly
transmitted through the data streams. For pointers, the corresponding
object is typically a distributed object as well, and its identifier
and potentially version is transmitted in place of its pointer.

@@

First, it configures the objects to retain its data \textsf{latency+1} commits,
which corresponds to the typical use case where objects are committed once per
frame. This allows render clients, which often are behind the application
process, to map objects with an old version. This does not necessarily translate
into increased memory usage, since new versions are only created when the object
was dirty during commit.

@@

The event handling code reproduced here is just showing the handling of one
type of event. A detailed description on how to customize event handling can be
found in \sref{sEventHandling}:

@@

{\footnotesize\begin{lstlisting}
void eq::Channel::frameDraw( const uint128_t& frameID )
{
    applyBuffer();
    applyViewport();

@@

The actual rendering uses display lists or vertex buffer objects. These
OpenGL objects are allocated using the object manager. The rendering is
done by the leaf nodes, which are small enough to store the vertex
indices in a \textsf{short} value for optimal performance with VBOs.
The leaf nodes reuse the objects stored in the object manager, or create
and set up new objects if it was not yet set up. Since one object
manager is used per thread (pipe), this allows a thread-safe sharing of
the compiled display lists or VBOs across all windows of a pipe.

@@

Carbon events issued for AGL windows are initially receive by the node
main thread, and automatically dispatched to the pipe thread. The pipe
thread will dispatch the event to the appropriate event handler. One
\textsf{AGLEventHandler} per window is used, which is created during
\textsf{AGLWindow::configInit}. The event handler installs a Carbon
event handler for all important events. The event handler uses an
\textsf{AGLWindowEvent} to pass the Carbon \textsf{EventRef} to
\textsf{AGLWindowIF::processEvent}. During window exit, the installed
Carbon handler is removed when the window's event handler is deleted. No
event handling is set up for pbuffers.

For each GLX window, one \textsf{GLXEventHandler} is allocated. FBOs and
pbuffers are handled in the same way as window drawables. The event dispatch
finds the corresponding event handler for each received event, which is then
used to process the event in a fashion similar to AGL. The
\textsf{GLXWindowEvent} passes the \textsf{XEvent} to
\textsf{GLXWindowIF::\-process\-Event}.

Each \textsf{WGLWindow} allocates one \textsf{WGLEventHandler} when the
window handle is set. The WGLEventHandler passes the native event
parameters \textsf{uMsg}, \textsf{wParam} and \textsf{lParam} to
\textsf{WGL\-Win\-dowIF::processEvent} as part of the
\textsf{WGLWindowEvent}. No event handling is set up for pbuffers.

@@

This section explains window initialization in detail. It discusses in
detail the handling of the different window systems. Window systems are
integrated through the \textsf{SystemWindow} interface, which externalizes
the details of the windowing API from the \textsf{eq::Window}
implementation and facilitates the integration with new windowing APIs
and custom applications.

@@

An application typically chooses to subclass the sample implementation if only
minor tweaks are needed for integration. For major changes or new window
systems, it is often easier to subclass directly from \textsf{SystemWindow} or
\textsf{GLWindow} and implement the abstract methods of this interface class.

@@

\begin{enumerate}
\item\textsf{initWGLAffinityDC} is used to set up an affinity device
  context, which might be needed for window creation. The WGL window
  tracks potentially two device context handles, one for OpenGL context
  creation (the affinity DC), and one for \textsf{swapBuffers} (the
  window's DC).
\item\textsf{chooseWGLPixelFormat} chooses a pixel format based on the
  window attributes. If no device context is given, it uses the system
  device context. The chosen pixel format is set on the passed device
  context.
\item\textsf{configInitWGLDrawable} creates the drawable. The device
  context passed to \textsf{configInitWGLDrawable} is used to query the
  pixel format and is used as the device context for creating a
  pbuffer. If no device context is given, the display device context is
  used. On success, it sets the window handle. Setting a window handle
  also sets the window's device context.
\item\textsf{createWGLContext} creates an OpenGL rendering context using
  the given device context. If no device context is given, the window's
  device context is used. This function does not set the window's OpenGL
  context.
\end{enumerate}

@@

For head-mounted displays (HMD), the tracking information is used to move the
frusta with the observer, as shown in \fig{fHMD}. This results in different
projections compared to normal tracking with fixed projection screens. This
difference is transparent to Equalizer applications, only the configuration file
uses a different wall type for HMDs.

@@

The focus distance calculation relative to the observer is similar to the origin
algorithm, but it keeps the closest wall in the observer's view direction at the
given focus distance, as shown in \fig{fObserver}. When the observer moves
forward, the focal plane moves forward as well. Consequently, when the observer
looks in a different direction, a different object in the scene is focused, as
indicated by the dotted circle in \fig{fObserver}.

@@

On a high level, the compositor combines multiple input frames using 2D
tiling, depth-compositing for polygonal data or sorted, alpha-blended
compositing for semi-transparent volumetric data. These operations
composite either directly all images on the GPU, or use a CPU-based
compositor and then transfer the preintegrated result to the GPU. The
high-level entry points automatically select the best algorithm. The
CPU-based compositor uses, based on availability, OpenMP and the
Paracomp compositing library to accelerate its operation.

@@

Each \textsf{LocalNode} provides a \textsf{co::Zeroconf} communicator, which
allows node and resource discovery. This requires compiling using a Lunchbox
library supporting dns\_sd. The service "\_collage.\_tcp" is used to announce
the presence of a listening \textsf{LocalNode} using the ZeroConf protocol to
the network, unless the local node has no listening connections. The node
identifier and all listening connection descriptions are announced using keys
starting with "co\_". Internally Collage uses this information to connect
unknown nodes by using the node identifier alone.

Applications may use the ZeroConf communicator to add additional key-value pairs
to announce application-specific data, and to retrieve a snapshot of all
key-value pairs of all discovered nodes on the network.

@@

\textsf{co::Serializable} implements one typical usage pattern for data
distribution based on \textsf{co::Object}. The \textsf{co::Serializable} data
distribution is founded on the concept of dirty bits, allowing inheritance with
data distribution. Dirty bits form a 64-bit mask which marks which parts of the
object have to be distributed during the next commit. It is easier to use, but
imposes one typical way to implement data distribution.

@@

\begin{tabularx}{\textwidth}{|l|X|}
  \hline
  \textbf{Name} & EQ\_CONFIG\_IATTR\_FOCUS\_MODE\\
  \textbf{Value} & fixed \textbar relative\_to\_origin \textbar
  relative\_to\_observer\\
  \textbf{Default} & fixed\\
  \textbf{Details} & The mode how to use the focus distance. The
  focus mode and distance influence the calculation of stereo frusta. See
  \sref{sFocusDistance} for details.\\
  \textbf{See also} & EQ\_CONFIG\_IATTR\_FOCUS\_DISTANCE
  EQ\_CONFIG\_FATTR\_EYE\_BASE\\
  \hline
\end{tabularx}\\\vfill

@@

\begin{tabularx}{\textwidth}{|l|X|}
  \hline
  \textbf{Name} & EQ\_WINDOW\_IATTR\_HINT\_STEREO\\
  \textbf{Value} & OFF \textbar \ ON \textbar \ AUTO\\
  \textbf{Default} & AUTO\\
  \textbf{Details} & Determines if the window selects a quad-buffered
  stereo visual. When set to AUTO, the default window initialization methods try
  to allocate a stereo visual for windows, but fall back to a mono
  visual if allocation fails. For pbuffers, AUTO selects a mono visual.\\
  \textbf{See also} & EQ\_COMPOUND\_IATTR\_STEREO\_MODE\\
  \hline
\end{tabularx}\\\vfill

\begin{tabularx}{\textwidth}{|l|X|}
  \hline
  \textbf{Name} & EQ\_WINDOW\_IATTR\_HINT\_DOUBLEBUFFER\\
  \textbf{Value} & OFF \textbar \ ON \textbar \ AUTO\\
  \textbf{Default} & AUTO\\
  \textbf{Details} & Determines if the window selects a double-buffered
  stereo visual. When set to AUTO, the default window initialization methods try
  to allocate a double-buffered visual for windows, but fall back to a
  single-buffered visual if allocation fails. For pbuffers, AUTO selects
  a single-buffered visual.\\
  \hline
\end{tabularx}\\\vfill

@@

\begin{tabularx}{\textwidth}{|l|X|}
  \hline
  \textbf{Name} & EQ\_WINDOW\_IATTR\_HINT\_DRAWABLE\\
  \textbf{Value} & window \textbar \ pbuffer \textbar \ FBO \textbar \ OFF\\
  \textbf{Default} & window\\
  \textbf{Details} & Selects the window's drawable type. A window is an
  on-screen, window system-dependent window with a full-window OpenGL
  drawable. Pbuffers are off-screen drawables created using window
  system-dependent pbuffer APIs. FBO are off-screen frame buffer
  objects. A disabled drawable creates a system window without a frame buffer
  for context sharing, e.g., for asynchronous operations in a separate
  thread. To calculate the pbuffer or FBO size on unconnected devices, a pipe
  viewport size of 4096x4096 is assumed, unless specified otherwise using the
  pipe's viewport parameter.\\
  \hline
\end{tabularx}\\\vfill
